---
title: "Elections Exercise"
author: "Evin Smith"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    math: katex
---

A demonstration a couple different ways you could look at election data



```{r setup, include=FALSE}
rm(list=ls())
options(scipen = 999)
library(tidyverse)
library(elections)
library(randomForest)
data("eldat")
```

# Data Dictionary 
 
## American presidential election winners from 1920-2016

electionyear: Calendar year in which the election was held.

presel.Date: Date at which the election was held.

winner: Name of the winner.

winnerparty: Party of the winner.

winnerparty.tmin1: Party of the winner, one election earlier.

winnerparty.tmin2: Party of the winner, two elections earlier.

winnerparty.tmin3: Party of the winner, three elections earlier.

winnerparty.tmin4: Party of the winner, four elections earlier.
runnerup Name of the runner up.

runnerupparty: Party of the runner up.

popvotepercmargin.rep: Popular vote margin (%) of the republican party as compared to the
democratic party.

popvotepercmargin.rep.tmin1: Popular vote margin (%) of the republican party as compared to
the democratic party, one election earlier.

turnoutperc: Turnout (%).
turnoutperc.tmin1: Turnout (%), one election earlier.

djia.reldiff: The relative change (%) of the Dow Jones index in the four years before the election.

dispincome: Per capita disposable income (2009 dollars) in the calendar year of the election.

dispincchange: Relative change (%) of the per capita disposable income over the four years before
the election.

# Considerations 
## Limited data

Limited features on social and global factors, major world event such as the COVID-19 pandemic would be expected to have an impact on future outcomes.

# EDA

### Data Quality Issues to note 
Missing disposable income data  for earlier elections

```{r, warning=FALSE}
eldat %>% group_by(winnerparty) %>% count()
eldat %>% distinct(winnerparty,winner) %>% group_by(winnerparty) %>% count()

summary(eldat%>% select(dispincome,dispincchange,djia.reldiff,
                         turnoutperc.tmin1,turnoutperc,popvotepercmargin.rep,
                         popvotepercmargin.rep.tmin1))


summary(eldat%>%filter(winnerparty=="Rep.") %>%
          select(dispincome,dispincchange,djia.reldiff,,turnoutperc,popvotepercmargin.rep))

summary(eldat%>%filter(winnerparty=="Dem.")%>%
          select(dispincome,dispincchange,djia.reldiff,,turnoutperc,popvotepercmargin.rep))


 eldat %>% group_by(winner,winnerparty) %>%
   count() %>%
   ungroup() %>%
   group_by(number_of_terms=n,winnerparty) %>% 
   count()

```
### Key Findings 

When a republican is selected as a winner the median disposable income and the relative change (%) of the Dow Jones

index in the four years is higher, median voter turnout is lower. 

Of the 9 distinct republican candidates that have won between the years of 1920-2016 4 (45%) served more than 1 term

of the 7 distinct democrat candidates that have won between the years of 1920-2016 3 (43%) served more than 1 term,
with Franklin Roosevelt serving 4 terms from 1932-1944




```{r, warning=FALSE}
#Fit model for probability that the winner is a republican
elmod1 <- glm(winnerparty == "Rep." ~ dispincome+dispincchange+djia.reldiff+turnoutperc+
               popvotepercmargin.rep,
data = eldat, family = binomial(link = logit))
summary(elmod1)

elmod2 <- glm(winnerparty == "Rep." ~ winnerparty.tmin1*winnerparty.tmin2+winnerparty.tmin3+winnerparty.tmin4,
data = eldat, family = binomial(link = logit))
summary(elmod2)

elmod3 <- glm(winnerparty == "Dem." ~ dispincome+djia.reldiff+turnoutperc,
data = eldat, family = binomial(link = logit))

summary(elmod3)

eldat_rf <- eldat %>%mutate(winnerparty=case_when(winnerparty=="Dem."~1,
                                                  TRUE ~0),
                            winnerparty.tmin1=case_when(winnerparty=="Dem."~1,
                                                  TRUE ~0),
                            winnerparty.tmin2=case_when(winnerparty=="Dem."~1,
                                                  TRUE ~0))


m.rf<- randomForest(
  winnerparty ~ 
    winnerparty.tmin1+
    winnerparty.tmin1+
    turnoutperc+
    turnoutperc.tmin1+
    djia.reldiff,
    data = eldat_rf,
  importance = T,
  nperm = 2,
  na.action = na.omit,
  mtry = 3
)

importance(m.rf)
m.rf



```
### Key Findings 

The party that won two elections previously has a negative impact on who will win in the current election. 

If a party has been held the office for two terms they are less likely to win in the current election 

###  Other supplemental data idea

Include exchange rate data in the model to see how the countries currency does against other major currencies 

Include unemployment by age group, specifically broken down by generation like (baby-boomers, gen-z, millenals etc)

(Idea is to try to include features that are captured by age group)

Include demographic data from past elections 

Include housing rate data 

Include immigration data (how many people are moving to the country from another, how many citizens are leaving)

Birth rates in the country 







